---
title: "Reproducible Reasearch Course Project 1"
author: "Tim Knox"
date: "January 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instruction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

1) Load the data (i.e. read.csv())
2) Process/transform the data (if necessary) into a format suitable for your analysis


```{r}
setwd("C:/Users/Timothy.Knox/Desktop/Coursera/Reproducible Research/Course Project 1")
activity_data <- read.csv("./activity.csv")
```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1) Calculate the total number of steps taken per day.
```{r}
##Remove NA's from Steps columns
na_removed <- activity_data[!is.na(activity_data$steps), ]

##Aggregate for sum of steps for each day
aggregate_data <- aggregate(steps ~ date, na_removed, sum)
```


2) Make a histogram of the total number of steps taken each day.
```{r}
##Create Histogram with number steps on x-axis and number of days on y-axis
hist(aggregate_data$steps, breaks = 7, xlab = "Number of Steps", ylab = "Frequency of Days",
    main = "Total Steps per Day", col = "green")
```

3) Calculate and report the mean and median of the total number of steps taken per day.

```{r}
##Output the mean of the total number of steps taken each day
mean(aggregate_data$steps)

##Output the median of the total number of steps taken each day
median(aggregate_data$steps)
```

## What is the average daily activity pattern?

1) Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
```{r}
##Aggregate data by time intervals, average number of steps, and across all days
aggregate_data <- aggregate(steps ~ interval, na_removed, mean)

##Plot intervals(x axis) against number of steps (y axis) for each day
plot(aggregate_data$interval, aggregate_data$steps, type = "l", xlab = "5 Min Intervals",
     ylab = "Average Number of Steps", main = "Average Number of Steps per 5 Min Interval",
     col = "blue")
```

2) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
##Which 5min interval contains the maximum average number of steps
maxSteps <- max(aggregate_data$steps)
aggregate_data[aggregate_data$steps == maxSteps, 1]
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1) Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
```{r}
##Create new dataset that is equal to activity_data (and will later will have NA values replaced)
activityData2 <- activity_data
##Vector with all NA Rows
naRows <- is.na(activityData2$steps)
```

2) Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}
##Calculate mean for each 5min Interval
avgInterval <- tapply(activityData2$steps, activityData2$interval, mean, na.rm = TRUE, simplify = TRUE)
```

3) Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
##Replace NA Rows with the Interval average
activityData2$steps[naRows] <- avgInterval[as.character(activityData2$interval[naRows])]
```

4) Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
##Create histogram of total number of steps taken each day
totalSteps <- aggregate(steps ~ date, activityData2, sum)
hist(totalSteps$steps, breaks = 10, xlab = "Number of Steps", ylab = "Frequency of Days", 
     main = "Total steps per Day", col = "green")

##Calculate Mean and Median total number of steps each day
mean(totalSteps$steps)
median(totalSteps$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1) Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
##Create a new factor variable in the dataset with two levels - "weekday" and "weekend" 
##  indicating whether a given date is a weekday or weekend day
activityData2$date <- as.Date(activityData2$date)
weekdays <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activityData2$weektype <- c('Weekend', 'Weekday')[(weekdays(activityData2$date) %in% weekdays) + 1L]
```

2) Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
library(ggplot2)
weektypeAvg <- aggregate(steps ~ interval + weektype, activityData2, sum)
ggplot(weektypeAvg, aes(x = interval, y = steps, color = weektype)) +
  geom_line() +
  ggtitle("Average Daily Steps by Weektype") +
  xlab("Interval") +
  ylab("Number of Steps") +
  facet_wrap(~weektype, ncol = 1, nrow = 2)
```

